\section{Introduction}\label{sec:intro}

\subsection{Motivation}\label{sec:intro:motivation}

Research workflows in machine learning and adjacent fields share a
common skeleton: brainstorm a research idea, design and run an
experiment, analyse the result, write a paper, and iterate. Each step
is increasingly tractable for large language models, but stitching
them into a single autonomous loop has, so far, been an open
engineering problem. The published systems that come closest --- AIDE,
the AI~Scientist family, VirSci, and PaperBench evaluators --- each
fix one or two of these axes while leaving the others manual.

ARI exists to fill that gap. We assume the user supplies a research
question and a compute budget; the system returns a publishable paper,
the experimental code that backs every claim, and a complete audit
trail of how it got there. The audit trail is non-negotiable: every
node of the search tree, every tool call, every model invocation, and
every dollar of API spend is recorded inside a single \term{checkpoint
directory} (\cite{jiang2025aide} adopt a similar discipline). The
checkpoint is the unit of reproducibility.

\subsection{Contributions}\label{sec:intro:contrib}

This report formalises three contributions:

\begin{enumerate}
  \item A \textbf{best-first tree search} over research-step nodes
        whose selection score $\utility{n}$ separates the empirical
        reward, a novelty term, and a depth penalty
        (\cref{sec:exploration:formalization}).
  \item A \textbf{paper-generation pipeline} keyed off a hierarchical
        rubric $\Rubric$ whose leaf judges grade individual claims;
        the system score is $\paperscore{\paper}=\sum_i \rubricweight{i}\,\judgefn{i}(\paper)$
        (\cref{sec:paper:rubric}). The PaperBench replicator is
        integrated as a tool call (\cref{sec:paper:paperbench}).
  \item A \textbf{checkpoint scoping} discipline that makes the
        pipeline fully reproducible: every external state (memory,
        cost ledger, lineage pointers) is written below
        \texttt{\$ARI\_CHECKPOINT\_DIR}, never the user's home
        directory. This is the operational consequence of the
        \emph{determinism} design principle, one of five that
        \cref{sec:overview:principles} lays out in full
        (we refer to it as P2 there and in the rest of the report).
\end{enumerate}

\subsection{Scope of this report}\label{sec:intro:scope}

This report describes \texttt{ari-core} and the fourteen
\texttt{ari-skill-*} packages at version v0.8.0, which promotes the
PaperBench reproduction path to a protocol-faithful three-stage contract
--- agent rollout, reproduction, and grading --- adds host-truthful
environment guardrails that stop the rollout agent from assuming
capabilities the host does not provide, makes the search-evaluation
layers configurable, and carries out a first operational reproduction
pass on a real lossy-compression target (\cref{chapter:eval}). The body
describes the algorithms and data flow at the module level; the verbatim
LLM prompts the system uses are listed in \cref{sec:prompts}. The
empirical study is preliminary at this version --- \cref{chapter:eval}
reports the observations available so far, and a controlled benchmark is
left to future work.

\subsection{Organisation}\label{sec:intro:organisation}

\Cref{sec:overview} gives a top-down view of ARI: the orchestrator,
the fourteen skills, the pipeline DAG, and the design principles.
\Cref{sec:exploration,sec:paper} are the two technical cores, each
with its own formal definitions and empirical observations.
\Cref{chapter:eval} reports preliminary observations; \cref{sec:related}
positions the work; \cref{sec:discussion} collects open problems.
