\section{Preliminary Observations}\label{chapter:eval}

This chapter reports what the running system establishes at v0.9.0.
The evidence is of three kinds: an end-to-end generation run whose
released paper passed every verification gate
(\cref{sec:eval:sample}); properties that hold by construction and
are confirmed in operation (\cref{sec:eval:construction}); and an
earlier reproduction exercise carried out on the then-current v0.8.0
pipeline (\cref{sec:eval:repro}). All of it is preliminary: the
end-to-end observations are single runs rather than a controlled
benchmark, and the panel-scale study they call for is deferred
(\cref{sec:eval:limits}). Quantitative figures elsewhere in this
report are illustrative, generated from fixed-seed sample data so
that the document build is reproducible.

\subsection{End-to-end generation: a gate-verified sample paper}\label{sec:eval:sample}

One end-to-end result is demonstrated and inspectable: the released
sample paper, an eight-page study produced autonomously by a real
run. It studies compressed sparse row (CSR) sparse-dense matrix
multiplication (SpMM), selecting a store policy across
right-hand-side (RHS) widths under the guidance of a performance
model (the paper's ``loopline'' model) fitted from the run's own
measurements. Four observations make the run an existence proof for
the verification machinery this report describes.

First, all twelve falsifiable claims declared in the run's metric
contract --- minted once at idea time and returned verbatim for the
rest of the run (\cref{sec:paper:contract}) --- carried
machine-verified experiment evidence: the final claim--evidence gate
reported zero errors, so no number in the released paper outran what
the gate could re-derive from recorded results. Second, the
computed-evidence chain behind the paper's central selection ---
parameter fitting, then held-out validation of the fit, then
model-based selection among configurations --- executed through
lineage-chained nodes (\cref{sec:exploration:lineage-chaining})
rather than collapsing into repeated probes, so the failure mode
that mechanism was introduced to close did not recur. Third,
correctness was validated within the run against an independent
dense single-precision (FP32) reference, with a maximum absolute
error of $6.79\times10^{-6}$. Fourth, the semantic review flagged
five overclaims in the draft; after the refine pass all five were
resolved and none remained --- and because the resolved count is a
raw before/after delta (\cref{sec:paper:contract}), a refine that
had introduced new overclaims would have surfaced as a regression.

We read this result narrowly. Its value is the \emph{verified}
property --- every quantitative claim in the released paper is bound
to a recorded artefact and passed the deterministic gate --- not any
headline number the paper reports; the paper itself hedges
explicitly wherever a statement outruns its artefact grounding. The
property is worth stating because earlier autonomous pipelines lack
it: in the closest published predecessor, hallucination control is
prompt-level, the automated review of a finished draft is advisory
rather than blocking, generated papers were observed to hallucinate
experimental details, and the authors list automatic verification of
reported results as future work~\cite{lu2024ai}. A single verified
run demonstrates that the contract, lineage chaining, gate, and
review--refine loop compose end to end; it does not establish a
distribution, which is exactly the controlled study deferred below.

\subsection{Properties holding by construction}\label{sec:eval:construction}

Three further properties hold by construction and are confirmed in
operation. Cost accounting is exact: a dollar cost is attached to
every model invocation and accumulated, per run, into the
checkpoint's results record. Exploration termination is not
budget-only: the stagnation detector
(\cref{sec:exploration:termination}) provides a content-based exit
and has been observed to fire before the step and cost budgets
($\Tmax$ / $\Cmax$) are exhausted. And the execution-profile pathway
behaves as specified (\cref{sec:paper:paperbench:v072}): rubric
generation emits a multi-rank Message Passing Interface (MPI)
execution profile, with the paper's reported scale envelope, for a
high-performance-computing (HPC) paper and omits it for a
single-machine paper, while a multi-flag cluster dispatch is
accepted by the local scheduler once the runtime probes drop the
flags it does not support.

\subsection{An earlier reproduction exercise (v0.8.0)}\label{sec:eval:repro}

The system's first operational end-to-end exercise predates v0.9.0:
the then-current v0.8.0 pipeline was run against a published
lossy-compression target, a GPU error-bounded scientific compressor,
through the reproduction path of \cref{sec:paper:paperbench}. The
degree of reproduction achieved on this single target was partial
but substantive. The agent fetched a real slice of the paper's cited
simulation dataset rather than substituting synthetic data; compiled
native CUDA interpolation kernels and ran them on the GPU;
implemented Lorenzo and multi-level interpolation predictors with
per-level autotuning; and swept a range of error bounds, recovering
reconstruction qualities from roughly 50 to 90~dB peak
signal-to-noise ratio (PSNR) with the corresponding compression
ratios. Against the fine-grained rubric the run cleared on the order
of a tenth of the weighted leaves, for two reasons rooted in the
agent's behaviour rather than the pipeline's: the graded metric came
from a CPU predictor --- the GPU kernels were built and executed,
but the agent omitted their predicted quality from its reported
results --- and the agent terminated voluntarily after a small
fraction of its time budget, leaving the remaining datasets and the
proprietary lossless stage unattempted. Both behaviours match
failure modes documented for replication agents at large, where most
models end runs early claiming completion and score far higher on
writing code than on executing it and matching the paper's
results~\cite{starace2025paperbench}. The exercise predates the
verification machinery of \cref{sec:eval:sample} and is reported
unchanged, as the historical baseline.

\subsection{Limitations}\label{sec:eval:limits}

The evaluation remains preliminary. Neither result above is a
controlled benchmark: the generation run is a single trajectory, and
the reproduction exercise covers a single target paper. A controlled
evaluation --- median run cost, per-seed exploration curves, and
per-category rubric distributions over a frozen checkpoint, together
with full-rollout replication scores across a panel of contrasting
target papers --- is left to future work. The same study should
settle the empirical questions this report leaves open, such as
which termination cause (\cref{sec:exploration:termination}) fires
most often in practice.
