\section{Related Work}\label{sec:related}

We position ARI against four threads of prior art.

\subsection{Tree search for code}\label{sec:related:treesearch}

AIDE \cite{jiang2025aide} formalises ML-engineering as a tree search
over plan / code / debug states; AlphaCode \cite{li2022competition}
is the canonical reference for large-scale generate-and-filter on
competitive programming; the MLE-bench evaluator
\cite{chan2024mle} is now the standard for measuring
ML-engineering agents. ARI inherits the BFTS skeleton from AIDE but
adds (i) an LLM-driven candidate selector (\cref{sec:exploration:node-selection}),
which avoids fixing a closed-form scalar utility and lets the
selection prompt weigh \texttt{has\_real\_data}, full
\texttt{metrics}, depth, and a soft
\texttt{diversity\_bonus} jointly, and
(ii) the cross-run lineage (\cref{sec:exploration:lineage}) which
neither AIDE nor MLE-bench formalises.

\subsection{Autonomous research agents}\label{sec:related:autonomous}

The AI~Scientist family \cite{lu2024ai,lu2026towards} closes a similar
loop end-to-end (idea $\to$ experiment $\to$ paper) but does not
expose the rubric and refine cycle as a first-class object. VirSci
\cite{su2024many} simulates several scientist personas and uses
their consensus to score outputs; ARI can drive VirSci's original
deliberation engine directly and unmodified --- its freshness-based
\texttt{select\_coauthors} team formation and multi-agent
\texttt{generate\_idea} loop --- over a live Semantic~Scholar snapshot,
so that the multi-agent mechanism it builds on is the published one
executed against current literature rather than a paraphrase of it.
Agent~Laboratory
\cite{schmidgall2025agent} treats the agent as a collaborator
rather than a driver. ARI differs in two operational respects:
\textbf{everything lives in a checkpoint} (P1 / P4), and the rubric
is the contract between explore and write.

\subsection{Paper evaluation}\label{sec:related:eval}

PaperBench \cite{starace2025paperbench} is the closest analogue to our
\texttt{paper-re} integration; it pioneered the rubric-as-tree
formalism. We adopt the same per-leaf grade and aggregate weight
(\cref{eq:paperscore}) but extend it with a per-axis floor used as a
refine guard (\cref{sec:paper:failures}).

Story2Proposal \cite{qian2026story2proposal} governs structured manuscript
generation through a persistent shared contract --- section structure, a
registry of visual artifacts, and document-wide validation rules --- with
writer, refiner, and renderer agents coordinating under a
generate--evaluate--adapt loop scored on reasoning verification, data
fidelity, and visual coherence. It takes the experimental evidence as given and
governs how that evidence is written up; it does not execute experiments. ARI
integrates that contract-governed-generation principle and its evaluation axes
into an execution-grounded pipeline (\cref{sec:paper:evidence}): the contract is
realised by extending the structured experiment record with claims and numeric
assertions that reference stable node, result, and figure identifiers; data
fidelity becomes the deterministic claim--evidence gate that re-derives each
reported number from the recorded results; and reasoning and visual coherence
become a non-blocking evidence-grounded review that leaves the independent
manuscript review intact. Because ARI executes the experiments rather than
receiving the evidence as input, it extends this manuscript-level contract to
execution-grounded provenance.

\subsection{Foundations}\label{sec:related:foundations}

The agent loop follows the ReAct pattern \cite{yao2023react};
the refine cycle is descended from Self-Refine
\cite{madaan2023self} and Reflexion \cite{shinn2023reflexion};
and the choice to give the LLM a step-by-step scratchpad in
\emph{Think} is descended from Chain-of-Thought
\cite{wei2022chain}. None of these references propose a
checkpoint-scoped state; that is ARI's contribution.

An earlier draft cited a separate ML-research-agent benchmark, but we
removed that reference once it became clear that none of the candidate
sources met our verification rule (\cref{sec:paper:failures}); we keep
the related-work coverage to entries that can be S2-verified.
