\section{Discussion and Limitations}\label{sec:discussion}

\subsection{Determinism vs novelty}\label{sec:discussion:p2}

P2, the determinism principle, is the binding constraint of the
system. It constrains not whether the research memory
(\cref{sec:exploration:memory}) may use an embedding model but
\emph{where} the resulting non-determinism may act: the memory makes
no generative language-model calls, and its embedding-ranked archival
search --- dependent on a remote service, hence not byte-reproducible
--- stays off the reproducible path as an optional recall aid the loop
never pulls, unable by itself to ground a paper claim. The memory
reads the loop does depend on --- the working context injected at each node, the
verified context that grounds the paper --- are deterministic,
ancestor-scoped folds of per-node reports. Every reproducibility win
closes such a door to non-determinism; we accept the cost because the
checkpoint, not the run, is the contract with downstream users.

\subsection{Instruction-following at structural seams}\label{sec:discussion:seams}

The recurring lesson of v0.9.0 is that instruction-following degrades
at structural seams --- an index to parse, a name to reuse, a number
to declare --- and that the durable remedy is parser-side absorption,
not prompting. An unparseable selection index falls back to a
deterministic ranking (\cref{sec:exploration:fallback}); a model asked
twice names the same quantity differently, so the metric contract is
minted once and replayed verbatim; the writer's recurring
numeric-declaration quirks are normalised at parse time
(\cref{sec:paper:contract}). The catalogue of absorbed quirks is
reactive, grown from real runs, so an unanticipated variant first surfaces as a gate
mismatch --- conservatively so: it is flagged, an unresolved mismatch
blocks the final check, and nothing is silently accepted.

The same discipline fixes what may block. The gate certifies that
every quantitative claim matches a recorded artefact, while the
semantic review's subjective findings steer the refiner
(\cref{sec:paper:loop}) but never gate publication --- unlike the
wholly post-hoc review of \cite{lu2024ai} --- because a subjective
verdict on the blocking path would be an unauditable veto. A verified
paper can therefore still be a dull one.

\subsection{Scaling limits}\label{sec:discussion:scaling}

The single-machine assumption underlying the BFTS run-loop is the
biggest scaling constraint: a single \texttt{ThreadPoolExecutor}
(\cref{sec:exploration:run-loop}) is the sole worker pool, capped at
four workers. Multi-host expansion would require extending the
explicit-lineage discipline (P4; \cref{sec:overview:principles}) to
merges, not just ancestor walks --- the same gap that bounds computed
evidence, whose chains (\cref{sec:exploration:lineage-chaining}) are
scheduled today by name hints along a single lineage rather than by an
explicit scheduler over the evidence dependency graph. A separate
constraint is the LLM grader, empirically a dominant cost item
alongside the explorer; caching its output by (paper-hash,
rubric-hash) would help and has not yet shipped.

\subsection{Outlook}\label{sec:discussion:outlook}

The widest gap is evaluation breadth: \cref{chapter:eval} rests on one
reproduction exercise (run on the then-current v0.8.0 pipeline) and
one verified generated paper --- evidence that the machinery composes,
not a distribution. Future work is threefold: a controlled study over
a frozen checkpoint and a panel of contrasting target papers, in the
protocol spirit of \cite{chan2024mle,starace2025paperbench};
variance-aware aggregation over repeated rollouts and grader passes in
place of single-scalar roll-ups, the judges themselves being noisy;
and broader domains --- widening the evidence kinds the contract can
name beyond scalar measurements, as the execution-profile contract
(\cref{sec:paper:paperbench:v072}) widened the computations ARI can
host, toward fields whose primary evidence is not a single metric.
